Impact of Demographics and Skills of the Observer on Visual Statistical Inference

### 2013-06-02 sociology_paper.tex
irb and nsf note
tower example

\subsection{Visual Statistical Inference} The fundamental concept of visual statistical inference is introduced by \citet{buja:2009}. In their paper, they proposed lineup protocol to discover significant pattern in the data. Later these concepts are further developed by defining necessary terminologies and validating using a head to head comparison with conventional inference by \citet{majumder:2013}. The power of the visual test is obtained and it is observed that the power is as good as that of conventional test while in some scenarios, the power exceeds the conventional test power.

The test statistic used in visual inference is a plot of the observed data. This plot, called actual data plot, is then placed on a layout of plot called lineup. The rest of the plots in a lineup are generated from the model specified by the null hypothesis. An observer is then asked to evaluate the lineup. If the observer can detect the actual data plot in the lineup, the hypothesis is rejected. Figure \ref{fig:lineup_geno} displays a lineup of 20 plots where one of the plots is observed data. The rest of the 19 plots are generated from the assumption that there is no difference between two groups of data. For this, the group structure of the data is broken by random permutation of the groups. If the null hypothesis is true, the observed data should look similar to the rest of the plots in the lineup making it difficult to detect the observed plot. On the other hand if the null hypothesis is not true, observed data should be different from the rest of the plots in the lineup.

@@

\begin{figure}[htbp] 
   \centering
   \includegraphics[width=6.5in]{plot_turk9_geno_4_3.png} 
   \caption{A lineup of 20 plots. One of the plots is observed data and rest of the 19 plots are obtained by randomly permuting the group structure. Can you identify the observed data?}
   \label{fig:lineup_geno}
\end{figure}

@@

>>>>

\subsection{Visual Statistical Inference} The fundamental concept of visual statistical inference is introduced by \citet{buja:2009}. In their paper, they proposed lineup protocol to discover significant pattern in the data. Later these concepts are further developed by defining necessary terminologies and validating using a head to head comparison with conventional inference by \citet{majumder:2013}. The power of the visual test is obtained and it is observed that the power is as good as that of conventional test while in some scenarios, visual power exceeds that of conventional tests.

The test statistic used in visual inference is a plot of the observed data. This plot, called actual data plot, is then placed on a layout of plot called lineup. The rest of the plots in a lineup are generated from the model specified by the null hypothesis. An observer is then asked to evaluate the lineup. If the observer can detect the actual data plot in the lineup, the hypothesis is rejected. Figure~\ref{fig:tower} displays a lineup of 20 plots where one of the plots is observed data, while the remaining 19 plots are rendered from data generated under a null model. Which of these towers is the most extreme? Which side is higher? In an experiment XXX of XXX observers picked plot \# XXX as the most extreme plot from the lineup, and all agreed that the blue side was higher. 
This gives us a visual $p$-value of XXX, indicating that we have enough evidence to reject the null hypothesis. 
What does this mean, though? For that, we need to know the context of the data and we need to have more information about the generation of the null plots: in this example, we are investigating, whether the results from the 2012 US presidential election is consistent with polling results. As a null hypothesis we  assume that  the election results are consistent with polling results. The polling results give us a null model to generate data. Each poll gives us a margin between candidates. This is what we use as the mean of a normal distribution. In the absence of information of error for a poll, we will assume a standard deviation of 3 percentage points for all of the polls. While this is a severe simplification, the resulting margin of error of $\pm 6$ percentage points is realistic, and could be easily replaced if additional information on a poll became available. A null data set is then generate as a set of draws from normal distributions based on the latest poll results. Each panel in figure~\ref{fig:tower} shows an `electoral building' \citet{nytimes}: each state is represented by a rectangle. The width of a rectangle is determined by the difference between (polling) results, its height is given by the number of electoral votes, and fill color shows party affiliation (blue for Democrat, and red for Republican). Rectangles are placed on top of each other, separated by affiliation and sorted by margin between polls. A party needs 270 electoral votes or more to win the presidential election.

%The rest of the 19 plots are generated from the assumption that there is no difference between two groups of data. For this, the group structure of the data is broken by random permutation of the groups.

If the null hypothesis is true, the observed data should look similar to the rest of the plots in the lineup making it difficult to detect the observed plot. On the other hand, if the null hypothesis is not true, the observed data should be different from the rest of the plots in the lineup and observers should be able to pick the data plot out from the rest.

@@

%\begin{figure}[htbp] 
%   \centering
%   \includegraphics[width=6.5in]{plot_turk9_geno_4_3.png} 
%   \caption{A lineup of 20 plots. One of the plots is observed data and rest of the 19 plots are obtained by randomly permuting the group structure. Can you identify the observed data?}
%   \label{fig:lineup_geno}
%\end{figure}

\begin{figure}[htbp] 
   \centering
   \includegraphics[width=0.9\textwidth]{tower.pdf} 
   \caption{ \label{fig:tower} Which of these towers is the most extreme? Which side is higher?}
\end{figure}

@@

%------------------------------------------------------------------------------------
\paragraph{Acknowledgments}
%------------------------------------------------------------------------------------
This work was funded in part by National Science Foundation grant DMS 1007697. All studies were conducted with approval from the internal review board IRB 10-347.
